---
title: "COVID-19 Regional Monthly Report"
author: "Clinton Tedja"
date: "A fully automated and reproducible report. `r format(Sys.time(), '%B %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    css: "styles.css"
---


<style>
body, h1, h2, h3, h4 {
    font-family: "Open Sans", sans-serif;
}

body {
    color: #031c2d;
}
h1 {
    color: #085387;
}
a, a:hover {
    color: #c2dbec;
}
pre {
    font-size: 10px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r data, include = FALSE}
library(tidyverse)
library(data.table)
library(lubridate)
library(knitr)
library(kableExtra)
library(sf)
library(rnaturalearth)
library(plotly)


# this time-series dataset, of confirmed cases is updated daily
# by John Hopkins University
raw_jhu_conf <- read.csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"))
raw_jhu_reco <- read.csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv"))
raw_jhu_deat <- read.csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"))

# I'm also creating here a quick function to enable simpler display of numerics in any text, plus removing scientific notation throughout
custom_number_format <- function(x){ifelse(x > 999999, paste(format(round((x/1000000), 1), nsmall=1, big.mark=","),"M"), format(round(x), nsmall=0, big.mark=","))}

options(scipen=10000)

# A quick note also that rmarkdown::render_site() is used here

```

```{r Tidying, include = FALSE}
# This report takes the following as countries of interest.
# You can easily swap out countries for those you are interested in to make up your regional report.

countries_in_region <- c("Afghanistan", "Bangladesh", "Bhutan",
                    "Cambodia", "India", "Indonesia", "Laos",
                    "Myanmar", "Malaysia", "Mongolia",
                    "Nepal", "Philippines", "Pakistan", "Vietnam",
                    "Sri Lanka", "Timor-Leste", "Thailand",
                    "Maldives", "Fiji", "Kiribati", "Marshall Islands",
                    "Micronesia", "Nauru", "Palau", "Papua New Guinea",
                    "Samoa", "Solomon Islands", "Tonga", "Vanuatu")

# We'll first define 'case_region' as all confirmed cases for select countries
#   NB: tidying here to fix dates are due to dates coming in with:
#   [1] wide format; [2] 'X' in dates; [3] and non-standard dates

case_region <- (raw_jhu_conf %>% 
  pivot_longer(names_to = "date", 
               values_to = "cases", 
               cols = -c(Province.State, Country.Region, Lat, Long)) %>%
  mutate(date = as.Date(str_remove(date, "X"), 
                        format = "%m.%d.%y")) %>%
  rename(country = Country.Region) %>%
  # then, since i'm only interested in my organization's regional operations
  filter(country %in% countries_in_region)) %>%
  # we'll tidy the recovered df before joining
  left_join(
  (raw_jhu_reco %>% 
        pivot_longer(names_to = "date", 
                     values_to = "cases", 
                     cols = -c(Province.State, Country.Region, Lat, Long)) %>%
        mutate(date = as.Date(str_remove(date, "X"), 
                              format = "%m.%d.%y")) %>%
        rename(country = Country.Region, recovered = cases) %>%
        select(-Lat, -Long, -Province.State)), 
  by = c("country", "date")) %>%
  # and tidy the deaths df before joining
  left_join(
  (raw_jhu_deat %>% 
        pivot_longer(names_to = "date", 
                     values_to = "cases", 
                     cols = -c(Province.State, Country.Region, Lat, Long)) %>%
        mutate(date = as.Date(str_remove(date, "X"), 
                              format = "%m.%d.%y")) %>%
        rename(country = Country.Region, deaths = cases) %>%
        select(-Lat, -Long, -Province.State)), 
     by = c("country", "date"))

# Then, seeing that we only have a cumulative count, let's add a daily increase
case_region <- case_region %>% 
  group_by(country) %>% 
  mutate(new_cases = c(cases[1],diff(cases)),
         new_recov = c(recovered[1],diff(recovered)),
         new_death = c(deaths[1],diff(deaths)))


```

```{r cumulative counts, include = FALSE}
# This df here is an aggregate number of cumulative cases at the end of each month
cases_month_end_agg <- case_region %>%
  group_by(date) %>% 
  summarize(agg_cases = sum(cases)) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month)%>%
  summarize(max = max(agg_cases))%>%
  arrange(max)

# This df here is the number of cumulative cases at the end of each month, disaggregated by country
cases_month_end_countries <- case_region %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month, country)%>%
  summarize(max = max(cases))%>%
  arrange(max)

# And an accompanying df pulling just those values at the end of last month, to build our narrative.
cases_month_end_countries_last <- cases_month_end_countries %>%
  filter(month == floor_date(Sys.Date(), "month") - months(1))



```



# Overview
The number of cumulative COVID-19 cases in operational and oversight countries in the region at the end of `r format(floor_date(Sys.Date(), "month") - months(1), format = "%B %Y")`, stood at `r custom_number_format(nth(cases_month_end_agg$max, -2))`. This is an increase of `r custom_number_format(nth(cases_month_end_agg$max, -2) - nth(cases_month_end_agg$max, -3))` compared with the end of `r (format(floor_date(Sys.Date()) - months(2), format = "%B %Y"))`. The two countries with the highest number of cases were `r cases_month_end_countries_last[match(nth(cases_month_end_countries_last$max, -1), cases_month_end_countries_last$max), 2]` (`r custom_number_format(nth(cases_month_end_countries_last$max, -1))` cases) followed by `r cases_month_end_countries_last[match(nth(cases_month_end_countries_last$max, -2), cases_month_end_countries_last$max), 2]` (`r custom_number_format(nth(cases_month_end_countries_last$max, -2))`). 


```{r Building the Map, echo = FALSE, results = 'HIDE',message = FALSE, warning = FALSE}

# Here we'll use the rnaturalearth, sf and plotly packages to create a map.
cases_asia_map <- ne_countries(scale = "medium", returnclass = "sf") %>%
  right_join(cases_month_end_countries_last,
             by = c("sovereignt" = "country")) %>%
  select(max, name, geometry)%>%
  na.omit()

cases_asia_map <- cbind(cases_asia_map, st_coordinates(st_centroid(cases_asia_map)))


# And the resulting plot:
ggplotly(
  (ggplot(cases_asia_map) +
    geom_sf(fill = "#dfe3e6", color = "#ffffff") +
    geom_point(aes(x = X, y = Y, size = max, name = name, text = paste("In", name, "there were", custom_number_format(max), "cumulative cases at the end of last month.")), color = "#85b7da") +
    coord_sf(xlim = c(60,180), expand = FALSE) +
    theme_void() + 
    theme(
      legend.position = "none",
      axis.line = element_blank()) +
    scale_size(range = c(0, 12)) +
    labs(caption = 'Cumulative case count at the end of last month')),
  tooltip = "text"
)

```


```{r Table, echo = FALSE}

# And using kable here, we build a simple table.


kable(
  (cases_month_end_countries_last %>%
    ungroup() %>%
    select(max, country) %>%
    arrange(desc(max)) %>%
    mutate(max = custom_number_format(max)) %>%
    plyr::rename(replace = c("max" = "total cumulative cases at the end of last month")))[,c(2,1)],
  # Then the table formatting with kableExtra
  format = "html", row.names = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(row = 0, color = "#85b7da") %>% 
  scroll_box(height = "300px")

```
  
<br/><br/>
<br/><br/>

# Changing Trends

``` {r Daily New Cases, include = FALSE}

# Now let's take a look at which months there had been a lot of daily increases in cases across the region, with 'new_avg_agg'

new_avg_agg <- case_region %>% 
    group_by(date) %>% 
    summarize(agg_new_cases = sum(new_cases),
              agg_new_recov = sum(new_recov),
              agg_new_death = sum(new_death)) %>%
    mutate(month = floor_date(date, "month")) %>% 
    group_by(month) %>% 
    summarize(avg_new_cases = mean(agg_new_cases),
              avg_new_recov = mean(agg_new_recov),
              avg_new_death = mean(agg_new_death)) %>%
     filter(month < max(month))

max(new_avg_agg$month)
summary(case_region$date)

# And here's a few quick conditional outputs to insert into the narrative, depending on whether last month was the worst or if the peak was at another time.
text_last_month <- custom_number_format(as.integer(new_avg_agg[match(nth(new_avg_agg$month, -1), new_avg_agg$month), 2]))

text_if_last_month_worst <- ifelse(new_avg_agg[match(nth(new_avg_agg$month, -1), new_avg_agg$month), 2] == max(new_avg_agg$avg_new_cases), "the worst month since the start of the pandemic.", "an improvement compared to the peak, in ")

text_worst_month <- ifelse(new_avg_agg[match(nth(new_avg_agg$month, -1), new_avg_agg$month), 2] == max(new_avg_agg$avg_new_cases), "", format(as.Date(as.integer(new_avg_agg[match(max(new_avg_agg$avg_new_cases), new_avg_agg$avg_new_cases), 1]), origin = "1970-01-01"), format = "%B %Y"))

text_worst_month_number <- ifelse(new_avg_agg[match(nth(new_avg_agg$month, -1), new_avg_agg$month), 2] == max(new_avg_agg$avg_new_cases), "", paste("(", custom_number_format(max(new_avg_agg$avg_new_cases)), "cases)"))


```



## Trends Over Time

Last month, there was an average of `r text_last_month` new cases daily. This is `r text_if_last_month_worst`  `r text_worst_month` `r text_worst_month_number`.


``` {r Daily New Cases Plot, echo = FALSE, results = 'HIDE',message = FALSE}

# Here we build an interactive plot using plot_ly
# First, we reshape the data slightly for our purposes, adding an 'All' to then cooperate with plot_ly's dropdown function

case_region_graph <- case_region %>% 
  group_by(date) %>% 
  summarize(new_cases = sum(new_cases)) %>% 
  cbind(country = c("All")) 

case_region_graph <- case_region %>% 
  rbind(case_region_graph) %>% 
  select(country, date, new_cases) %>% 
  arrange(factor(country, levels = c("All"))) %>%
  filter(date < floor_date(max(case_region$date), 'month'))


# Then, a quick function to enable the entry of multiple country dropdowns
#   this essentially puts it into a list that would work within the plotly
#   function, without having to retype the function a million times
cdd <- function(country_number){
  (list(
    method = "restyle", 
    args = list("transforms[0].value",
                unique(case_region_graph$country)[country_number]),
    label = unique(case_region_graph$country)[country_number]))}

# Then for the actual plot. Some extra guidelines: https://plotly-r.com/scatter-traces.html#lines
plot_ly(case_region_graph, 
        x = ~date, 
        y = ~new_cases, 
        type = 'scatter',
        size = I(0),
        mode = 'markers',
        line = list(width = 0.5, color = "#85b7da"),
        transforms = list(
                 list(
                   type = 'filter',
                   target = ~country,
                   operation = '=',
                   value = unique(case_region_graph$country)[1]
      ))) %>% layout(
        xaxis = list(title = "Country", zeroline = FALSE,showgrid = FALSE),
        yaxis = list(title = "New Cases", zeroline = FALSE,showgrid = FALSE),
        updatemenus = list(
           list(type = 'dropdown',
                active = 0,
  # This lapply function allows us to duplicate our newly created cdd function by the number of countries in the dataframe, without repeating the number many times
                buttons = lapply(1:(length(unique(case_region_graph$country))), cdd))))

```


## Last Month's Trends

``` {r Daily Cases by Country, include = FALSE}

# In the last code chunk, we were able to see aggregate trends in terms of new cases in the region. Now, let's disaggregate by country.
new_avg_country <- case_region %>% 
    mutate(month = floor_date(date, "month")) %>% 
    group_by(month, country) %>% 
    summarize(avg_new_cases = mean(new_cases),
              avg_new_recov = mean(new_recov),
              avg_new_death = mean(new_death)) %>%
    group_by(country)%>%
    # we can also add percentage changes over each month
    mutate(pct_change_case = ((avg_new_cases-lag(avg_new_cases))/
                           lag(avg_new_cases)*100)) %>%
    # as well as an additional column "month_change" demonstrating
    #   the difference in new cases for each country over the months
    group_by(country)%>%
    mutate(month_change = 
             avg_new_cases-lag(avg_new_cases, n=1, default = NA))
      
# it will also be useful to just see stats at the latest month
new_avg_country_latest <- new_avg_country %>%
    filter(month == floor_date(Sys.Date(), "month") - months(1))%>%
  arrange(desc(month_change))

new_avg_country_latest
custom_number_format(148.89631336)

```

In `r format(floor_date(Sys.Date(), "month") - months(1), format = "%B %Y")`, compared with `r format(floor_date(Sys.Date(), "month") - months(2), format = "%B %Y")` there was a notable increase of average daily new cases in  `r  new_avg_country_latest[match(max(new_avg_country_latest$month_change), new_avg_country_latest$month_change), 2]` (`r  custom_number_format(as.integer(new_avg_country_latest[match(max(new_avg_country_latest$month_change), new_avg_country_latest$month_change), 3]))` cases), and  `r  new_avg_country_latest[match(nth(new_avg_country_latest$month_change, 2), new_avg_country_latest$month_change), 2]` (`r  custom_number_format(as.integer(new_avg_country_latest[match(nth(new_avg_country_latest$month_change, 2), new_avg_country_latest$month_change), 3]))` cases). Meanwhile, there was a decrease in `r  new_avg_country_latest[match(min(new_avg_country_latest$month_change), new_avg_country_latest$month_change), 2]` (`r  custom_number_format(as.integer(new_avg_country_latest[match(min(new_avg_country_latest$month_change), new_avg_country_latest$month_change), 3]))` cases), and  `r  new_avg_country_latest[match(nth(new_avg_country_latest$month_change, -2), new_avg_country_latest$month_change), 2]` (`r  custom_number_format(as.integer(new_avg_country_latest[match(nth(new_avg_country_latest$month_change, -2), new_avg_country_latest$month_change), 3]))` cases).

```{r Last Month Trends, echo = FALSE}
# let's plot it to see the results for the latest months!
# This plot was useful in the early months of the pandemic, like May 2021
#   where it was important to see where changes at a month level.
#   But obviously, it became less relevant as the pandemic dragged on over many
#     months and with many waves.

ggplotly(
  (new_avg_country_latest %>%
    ungroup(country)%>%
    mutate(country = fct_reorder(country, month_change))%>%
             ggplot(aes(month_change, country, 
                        text = paste("In", country, "there was an average of\n", custom_number_format(avg_new_cases), "new daily cases in", format(month, format = "%B"), "which was an increase of\n", custom_number_format(month_change), "average new daily cases compared with the previous month."))) +
    geom_bar(stat = "identity", aes(fill = month_change>0))+
    theme_minimal()+
    labs(x = "Change in average daily new cases", 
         title = "Where did average daily cases increase/decrease?",
         subtitle = paste("between", (format(floor_date(Sys.Date()) - months(2), format = "%B %Y")),
         "and", (format(floor_date(Sys.Date()) - months(1), format = "%B %Y"))))+
    theme(legend.position = "none",
          plot.title=element_text(size=10),
          plot.subtitle=element_text(size=8),
          axis.title = element_blank())+
    scale_fill_manual(values=c("#85b7da", "#fcdc5d"))),
  tooltip = "text"
)


```